An Algorithm to Prune Integer Overflow Tests

### 2012-08-30 doc/Manuscript/CGO/paper.tex
Adding the section about the Instrumentation Framework

% What are integer overflows.
The most popular programming languages, including C, C++ and Java, limit the
size of primitive numeric types.
For instance, the \texttt{int} type, in Java, ranges from $-2^{31}-1$ to
$2^{31}$.
Consequently, there exist numbers that cannot be represented by these types.
In general, these programming languages resort to a {\em wrapping-arithmetics}
semantics~\cite{TODO} to perform integer operations.
If a number $n$ is too large to fit into a primitive data type $T$, then $N$'s
value wraps around, and $N \mbox{module} T_{max}$ wounds up represented
instead.
There are situations in which this semantics is acceptable~\cite{TODO}.
For instance, programers might rely on this behavior to implement hash functions
and random number generators.
On the other hand, there exist also situations in which this behavior might
lead a program to produce unexpected results.
As an example, in 1996, the Ariane 5 rocket was lost due to an arithmetic
overflow -- a bug that resulted in a loss of more than U\$370 million
dollars~\cite{Dowson97}.
% TODO We need to explain better the wrapping-arithmetics, and we need to cite it
% properly. What could be a good (official) citation?
% And we need to cite

@@

% A brief overview about integer overflows.
The $n$-bits two's complement arithmetics wraps around at multiples of
$2^n$.
As an example the representation of $16 \times 16$ in 8-bits two's complement
is $0$, because $0 \equiv (16\times16) \mod 2^8$.
Programmers can use this well-defined behavior in benign ways.
For instance, the wrapping arithmetics provides a cheap $\mod 2^n$ operator,
which is used in the implementation of hash-functions or random number
generators.
However, integer overflows can also be the sources of bugs that are hard to
identify.
As pointed by Dietz {\em et al.}~\cite{Dietz12}, integer overflows are the
sources of vulnerabilities in widely used software, such as OpenSSH and
Firefox.
These vulnerabilities could give an adversary the opportunity to execute
arbitrary code in the compromised application.

@@

@@

We have implemented our range analysis algorithm in LLVM 3.0, and have run
experiments on a Intel quad core CPU with a 2.40GHz clock, and 3.6GB of RAM.
Each core has a 4,096KB L1 cache.
We use Linux Ubuntu 10.04.4.
Our implementation of range analysis has 3,958 lines of
commented C++ code, our e/u-SSA conversion module has 672 lines, and our
instrumentation pass has XXX lines.
We have analyzed 428 C programs that constitute the LLVM test suite plus the
integer benchmarks in SPEC CPU 2006.
Together, these programs contain 4.76 million assembly instructions.
This section has two main goals.
First, we want to show that our range analysis is fast and precise.
Second, we want to demonstrate the effectiveness of our framework to detect
integer overflows.

@@

\noindent
\textbf{Precision.}
Our implementation of range analysis is remarkably precise, considering its
runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size in less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, we present a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are far from being trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark suite, which is publicly
available~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
We chose Stanford because these benchmarks do not read data from external
files; hence, imprecisions are due to library functions that we cannot
analyze.

>>>>

% What are integer overflows.
The most popular programming languages, including C, C++ and Java, limit the
size of primitive numeric types.
For instance, the \texttt{int} type, in Java, ranges from $-2^{31}-1$ to
$2^{31}$.
Consequently, there exist numbers that cannot be represented by these types.
In general, these programming languages resort to a {\em wrapping-arithmetics}
semantics~\cite{TODO} to perform integer operations.
If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
value wraps around, and $n \mbox{module} T_{max}$ wounds up represented
instead.
There are situations in which this semantics is acceptable~\cite{TODO}.
For instance, programers might rely on this behavior to implement hash functions
and random number generators.
On the other hand, there exist also situations in which this behavior might
lead a program to produce unexpected results.
As an example, in 1996, the Ariane 5 rocket was lost due to an arithmetic
overflow -- a bug that resulted in a loss of more than U\$370 million
dollars~\cite{Dowson97}.
% TODO We need to explain better the wrapping-arithmetics, and we need to cite it
% properly. What could be a good (official) citation?
% And we need to cite

@@

% A brief overview about integer overflows.
In addition to being a source of several bugs that are hard to identify, integer 
overflows can represent a serious security risk.
As pointed by Dietz {\em et al.}~\cite{Dietz12}, integer overflows are the
sources of vulnerabilities in widely used software, such as OpenSSH and
Firefox.
These vulnerabilities could give an adversary the opportunity to execute
arbitrary code in the compromised application. 
We will demonstrate this with an example.
The $n$-bits two's complement arithmetics wraps around at multiples of
$2^n$.
As an example the representation of $16 \times 16$ in 8-bits two's complement
is $0$, because $0 \equiv (16\times16) \mod 2^8$.
Programmers can use this well-defined behavior in benign ways.
For instance, the wrapping arithmetics provides a cheap $\mod 2^n$ operator,
which is used in the implementation of hash-functions or random number
generators.

@@

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass.
This means our library works on the LLVM IR\footnote{LLVM Intermediate Representation, available at \url{http://llvm.org/docs/LangRef.html}}
instead of working directly on the original source code. 
This fact brings the advantage of allowing us to rely on
other analysis, such as the range analysis, to check if it is safe to state that
the possible values of an instruction will always be within the range of
the declared type of the instruction. 
If we can state this, it means that it is impossible for an overflow to 
occur in the analyzed instruction, because the range analysis is a conservative analysis; 
in this case, we don't need to insert dynamic checks, and the instrumented program
will be smaller and run faster than it would run without this prunning.
%How our instrumentation works
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after that instructions. 
The LLVM IR has five instructions that may lead to an overflow: 
ADD, SUB, MUL, SHL, and TRUNC.

%ADD instrumentation
The ADD instruction performs the sum of two operands $op_1$ and $op_2$ and
stores the result in a variable $x$.
The semantics of an ADD instruction is $x = op_1 + op_2$. 
The goal of the instrumentation of the instruction $x$ is to ensure that after the
execution of the ADD instruction, $x$ stores the correct result of the sum,
without any data loss.
For this, we insert different checks for signed and unsigned instructions.
If we are dealing with an unsigned ADD instruction $x$, we insert the following check:
\code{if $x < op_1$ or $x < op_2$ then} handle overflow \code{else} continue to 
$x$'s next instruction. If, instead, we are instrumenting a signed instruction $x$, 
we will have the following assertion:
\code{if ($op_1 > 0$ and $op_2 > 0$ and $x < 0$) or ($op_1 < 0$ and $op_2 < 0$ and $x > 0$) then} 
handle overflow \code{else} continue to $x$'s next instruction.

%SUB instrumentation
The SUB instruction performs the subtraction of two operands $op_1$ and $op_2$ and
stores the result in a variable $x$.
The semantics of an SUB instruction is $x = op_1 - op_2$. 
Note that the SUB semantics is close to the ADD semantics. For this reason, the
instrumentation of the SUB instructions looks like the ADD instrumentation.
In this case, we insert different checks for signed and unsigned instructions, too.
If we are dealing with an unsigned SUB instruction $x$, we insert the following check:
\code{if $op_1 < op_2$ then} handle overflow \code{else} continue to 
$x$'s next instruction. Note that the assertion makes a single comparison. This is sound
because if $op_2$ is greater than $op_1$, the result would be negative. However, unsigned
integers can't represent negative numbers. In this case, we have a overflow.
The instrumentation of a signed SUB instruction $x$ will have the following assertion:
\code{if ($op_1 < 0$ and $op_2 > 0$ and $x > 0$) or ($op_1 > 0$ and $op_2 < 0$ and $x < 0$) then} 
handle overflow \code{else} continue to $x$'s next instruction.

%MUL instrumentation
The MUL instruction performs the multiplication of two operands $op_1$ and $op_2$ and
stores the result in a variable $x$.
The semantics of an SUB instruction is $x = op_1 \times{} op_2$. 
The dynamic check of the MUL instruction is the same for signed or unsigned
instructions, because it relies on DIV, the inverse operation. 
To verify if the overflow has occurred, we divide $x$ by $op_1$. If the result
is different of $op_2$, then we have an overflow to handle. 
In this case, we will have the following assertion: 
\code{if $op_1 \neq{} 0$ and $(x \div{} op_1) \neq{} op_2$ then} 
handle overflow \code{else} continue to $x$'s next instruction.
Note that if $op_1 = 0$, the result of the multiplication will never overflow,
because the result of the MUL will be $0$, too. In this case, if $op_1 \eq{} 0$,
we don't perform the DIV, to prevent a division by zero exception.
This technique of using the inverse operation to check for overflows doesn't work
for ADD and SUB instructions because the $n$-bits two's complement wrapping-arithmetics 
would give us false negatives in the assertions.

%SHL instrumentation
The SHL instruction performs the shift left operation. It receives a value $op_1$
and the number of bits $n$ the SHL will shift $op_1$ by; the operation's result
is stored in a variable $x$.
The semantics os a SHL instruction is $x = op_1 \times{} 2^n$. 
The SHL instructions result in overflows when the $op_1$ is positive and 
$x$ is negative or when $op_1$ is negative and $n \neq{} 0$, because 
the SHL in negative numbers would produce undefined results that are
as harmful as integer overflows. This assertion has the same semantics for
signed and unsigned instructions.
In this case, we will have the following check: 
\code{if ($op_1 > 0$ and $x < op_1$) or ($op_1 < 0$ and $n \neq{} 0$) then} 
handle overflow \code{else} continue to $x$'s next instruction.

%TRUNC instrumentation
The TRUNC instruction performs the truncation of the operand $op_1$
to a number $n$ of bits and stores the result in a variable $x$.
The semantics of a TRUNC instruction is $x = n$LessSignificantBits$(op_1)$.
The dynamic checking of a TRUNC instruction is done by expanding $x$ to
the datatype of $op_1$ and comparing the expanded $x$ with $op_1$. If they 
don't have the same value, some data has been lost by the TRUNC instruction.
In this case, we have the following check:
\code{if $cast(x, getDataType(op_1)) \neq{} op_1$ then} 
handle overflow \code{else} continue to $x$'s next instruction.
Note that the instrumentation of the TRUNC instructions is a very fragile point, 
because it would catch any truncation with loss of precision, even if it is a
benign one.
To avoid trouble and make it easier to our users, the instrumentation library 
can receive an argument to let the user decide whether to insert or not
instrumentation in the TRUNC instructions.

%Overview about the instrumentation technique
Our instrumentation library works by adding new instructions to
the target programs. The instrumentation of all the instructions works
in the same way: first, the pass inserts instructions to gather the
information needed to do the assertion.
Then, the pass inserts compare instructions to generate a boolean value
that indicates if there is an overflow or not. 
After this, the pass inserts a conditional branch: 
the false clause of the branch, where the control goes when there is no 
overflow and the program, leads to the first instruction after the instruction
which has received the instrumentation.
the true clause of the branch leads to a block where we do the
overflow handling.

%What we can do in the overflow handling
Our library has three options to handle the overflows. 
First option: Do nothing and ignore the overflows. This option
was used to do our experiments and verify the slowdown produced 
by the new instructions.
Second option: Give a message in stderr whenever an overflow occurs.
This is the standard option. The instrumented program remains with 
the same behavior, but whenever an overflow occurs, the program
prints \texttt{Overflow occurred in FileName.cpp, line X.} in the
standard error stream.
Third option: Abort the execution when an overflow occurs.
This is the more aggressive option. With this option enabled,
whenever an overflow occurs, the program is terminated, not allowing
the execution of the program with undefined behavior due to 
integer overflows.

@@

We have implemented our range analysis algorithm in LLVM 3.0, and have run
experiments on a Intel quad core CPU with a 2.40GHz clock, and 3.6GB of RAM.
Each core has a 4,096KB L1 cache.
We use Linux Ubuntu 10.04.4.
Our implementation of range analysis has 3,958 lines of
commented C++ code, our e/u-SSA conversion module has 672 lines, and our
instrumentation pass has 762 lines.
We have analyzed 428 C programs that constitute the LLVM test suite plus the
integer benchmarks in SPEC CPU 2006.
Together, these programs contain 4.76 million assembly instructions.
This section has two main goals.
First, we want to show that our range analysis is fast and precise.
Second, we want to demonstrate the effectiveness of our framework to detect
integer overflows.

@@

\noindent
\textbf{Precision.}
Our implementation of range analysis is remarkably precise, considering its
runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size in less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, we present a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are far from being trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark suite, which is publicly
available~\footnote{\url{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
We chose Stanford because these benchmarks do not read data from external
files; hence, imprecisions are due to library functions that we cannot
analyze.
